Natural Selection Across Three Generations of Americans

### 2024-02-21 response-to-reviewers-revision-1.qmd

*In a similar vein, it would be very welcome to reiterate reason why the
expected effect is stronger amongst low income etc people.*

We now detail the theory more in the abstract, and we've described the
substitution effects and income effects straight away in the intro.
We've tried to keep the language as simple and intuitive as possible.

*I was wondering why the authors control for PCAs in the PGI analyses.
As stated by the authors, this study is not a causal study but one of
correlations. The consequence of a correlation (natural selection)
should be independent of its causes.*

Note that controlling for principal components in a regression, *y \~
PGIs + PCs*, gives the same effect sizes as running an uncontrolled
regression on PGIs which have previously had the PCs residualized out,
<!--# TE not an expert here, but I think it should say y_resid here. I've residualised PGS but not rlrs, because it would complicate the interpetation of our beta under the Price equation. I think we should just remove this paragraph. We can explain our approach by saying the residualisation get's us closer to an unbiased PGS and cite others who have done the same thing.-->*y
\~ PGIs_resid*. We control rather than residualizing because this gives
the correct standard errors[^1], which might matter with our relatively
low N.

@@

[^1]: Because the controls are correctly accounted for when calculating
    degrees of freedom.

Why residualize or control for principal components at all? I think for
two reasons. First, we are ultimately interested in selection on the
"true PGI". Stripping out PCs is an admittedly imperfect way to control
for confounding population stratification. Second, PGIs are typically
residualized in this way, so by examining the effect of these we're
following the "industry standard" and learning about scores in the form
they are actually used.

To simplify interpretation, we've decided to residualise and then
standardize the PGIs on the principal components, instead of control for
them.

@@

More specifically, we're interested in the pattern of selection effects
across different PGIs, and how it correlates with the PGI's relationship
to income. So we want to include PGIs with small selection effects,
which we predict to have a low absolute correlation to income. Given the
sample size, these PGIs are very likely to have insignificant individual
selection effects, but we shouldn't ignore them in the cross-PGI
analyses.

When we examine relationships across scores, we indeed test for
significance, always by bootstrapping the sample.

*The authors state that the correlation between the PGI and fertility
quantifies natural selection. However, to my knowledge it is the
covariance following the Robert-Price Identity: Morrissey, M. B., L. E.
B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's
equation in observational studies of natural populations." Journal of
evolutionary biology 23.11 (2010): 2277-2288.*

We quantify natural selection with the regression coefficient of the PGI
on relative lifetime reproductive success. At times in the original
manuscript this was wrongly referred to as a correlation but thus has
now been updated. The polygenic scores are standardized to have a
standard deviation of 1. Under a univariate model, the regression
coefficient is the covariance divided by the variance in the independent
variable. The variance is standardized to 1, so the regression
coefficient is equal to the covariance and thus the two are
interchangeable.

In the previous manuscript we used a multivariate model, controlling for
genetic principal components. We now no longer control for the PCs. Now
we residualise the polygenic scores on the PCs before standardizing the
polygenic scores.

@@

# Reviewer 2

*The author sought to examine natural selection as indicated by
polygenic scores in a contemporary representative sample of the U.S. and
found that the results partially support the economic theory of
fertility as a result of natural selection among both white and black
individuals. The authors' hypotheses are based primarily on the economic
theory of fertility, which emphasizes a trade-off between the number of
children and income: i) higher income makes raising children costlier,
but ii) higher income makes having more children more affordable. As far
as I understand the results of the paper, the results support prediction
1, but are more ambiguous for prediction 2?*

*However, in my opinion the theoretical database, data and analysis are
adequate, the problem I have is that the MS is a bit confusing. It
should be made clearer why an analysis was done and what the results
mean in terms of rejecting or accepting the theoretical assumptions. The
possible link to biological selection should also be explained in more
detail, as this has been done in the case of education, income and
reproduction to some extent on the basis of comparable data (Kong et al.
2017, Beauchamp 2016, Fieder & Huber 2002) and also in many other
studies.*

We have added a sentence explicitly addressing our motivation in the
introduction. We have also clarified the biological theory in the Data
section. We provide an explanation of the intuition for our approach and
how it is established in biology, through the Robertson-Price identity,
to estimate the selection differential. TE TODO: Write a paragraph on
biol. selection. (Not sure what is meant here?)

### 2024-02-21 hrs-selection.qmd
Text edits for reviewers and to adjust for new results

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans. To preview our results, we confirm point 1 above, but only see partial and ambiguous support for point 2. (TODO: or can we be stronger?)

@@

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R *survey* package [@lumleysurvey2023].

@@

We estimate coefficients of PGS on RLRS. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, and self-rated health. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

Lastly, we would like to know natural selection's effect sizes. The bivariate correlation of PGS with RLRS gives the change in one generation in the PGS due to natural selection, measured in standard deviations. Polygenic scores contain error, so estimated correlations are biased towards zero compared to the correlation of the true PGS. They can be scaled up by

@@

where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource]. Moving from error correction in a univariate model to a multivariate model, controlling for principal components of genetic array data, requires a slightly more complex correction, although it uses the same parameters. To do this we use the error-corrected estimator developed by @becker2021resource.

SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository. The authors use GCTA to estimate heritability. When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead. We also perform error correction with twin heritabilites. To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^1]. We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power. This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2].

@@

[^2]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms. These were PV001-PV018 in the 2016 core survey. They estimated an $R^2_{PGS}$ of $0.003$. When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`. Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I. `r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`. Given the great uncertainty over the true correlation we are unable to justify using the error-corrected estimator with the ADHD polygenic score.

@fig-true-effects reports the error corrected estimates of selection. As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations. For twin-heritability, different relationships with fertility among variants not measured on the chip. Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.

Results from a few traits show substantial uncertainty regarding the true effect size. For asthma, when chip heritability is used, the effect is positive and significant, but negative and significant when twin heritability is used! Such a large disparity is because @becker2021resource estimate the chip heritability at 0.015 and we use a twin heritability 0f 0.55. For personality traits, chip heritability is often low (e.g. $<0.05$) but twin heritabilities are often much higher ($≈0.4$) implying substantial missing heritability. For these traits, the confidence intervals on twin heritabilities are enormous. Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.

@@

@@

corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
    list_rbind() |>
    filter(term %in% pgs) |>
    rename(pgs = term) |>
    left_join(correct_scores)

>>>>

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The motivation for our study is to establish the direction and magnitude of natural selection upon a range of traits in Americans and to test whether the economic theory can explain the selection. Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans. To preview our results, we confirm point 1 above, but only see partial and ambiguous support for point 2. (TODO: or can we be stronger?)

@@

The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection. However the approach is also based in quantitative genetic theory and is able to yield an approximate estimate of genetic change from one generation to the next.

The Robertson-Price Identity states that the change in a genetic trait between one generation and the next is equal to its covariance with relative fitness [@Price1970identity; @Robertson1966identity], assuming no mutations. In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness. As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score from one generation. We standardize our polygenic scores to have a mean of 0 and a standard deviation of 1. With this standardization the regression coefficient of RLRS on the PGS is equal to the covariance. Our approach to measuring natural selection is not novel, having been and detailed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Standardization of polygenic scores used estimates of the population mean and variance of the polygenic score, as estimated with sampling weights. Statistical tests are adjusted for clustering and stratification using the R *survey* package [@lumleysurvey2023].

@@

We estimate coefficients of PGS on RLRS. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, self-rated health and having ever smoked. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

Lastly, we would like to know natural selection's effect sizes. The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations. Polygenic scores contain error, so estimated effects are biased towards zero compared to the correlation of the true PGS. They can be scaled up by

@@

where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource]. SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository. The authors use GCTA to estimate heritability. When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead. We also perform error correction with twin heritabilites. To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^1]. We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power. This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2]. Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.

@@

[^2]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms. These were PV001-PV018 in the 2016 core survey. They estimated an $R^2_{PGS}$ of $0.003$. When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`. Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not very significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I. `r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`. Given the great uncertainty over the true correlation we were unable to justify using the error-corrected estimator with the ADHD polygenic score.

@fig-true-effects reports the error corrected estimates of selection. As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations. For twin-heritability, different relationships with fertility among variants not measured on the chip may also bias the estimates. Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.

<!--# REMINDER TO RE-ESTIMATE THE PHENOTYPIC EFFECT WHEN ALL THE CODE HAS BEEN REDONE. -->

@@

<!--# Put in some discussion of the fact the most robust findings seem show opposite change in the phenotype during the 20th century. Negative Flynn Effect... Environmental variation across time as surely more important.-->

@@

corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
    list_rbind() |>
    filter(term %in% pgs) |>
    rename(pgs = term) |>
    left_join(correct_scores, by = join_by(pgs))
